---
title: "Template"
author: "Studentnames and studentnumbers here" Marnix Peter Simon Lamers 2894404,Sophie Thijmen Thomas
date: "`r Sys.Date()`"
output: pdf_document
---

# Set-up your environment

```{r package_install, include=FALSE}
#install.packages("tidyverse")
#install.packages("yaml")
#install.packages("rmarkdown")
```

```{r packages}
require(tidyverse)
require(rmarkdown)
require(yaml)
```

# Title Page

Include your names Marnix Peter Simon Lamers,Sophie Thijmen Thomas

Include the tutorial group number 3.5

Include your tutorial lecturer's name Chantal Schouwenaar

# Part 1 - Identify a Social Problem

## 1.1 Describe the Social Problem

The problem that we will assess is particulate matter distribution across different regions in the Netherlands, and its correlation with health effects, more specifically, mortality rates. The datasets that we use measure PM2.5, this is particulate matter that is less than 2.5μm in size. Another frequently used measurement of particulate matter is PM10, which is particulate matter that is less than 10μm in size. We have chosen to use PM2.5 because its small size allows it to penetrate deeper into the lungs than larger particles (WHO, 2006).

A potential pitfall in our research is that Janssen et al. (2013) argues that both PM2.5 and PM10 have a very similar association with mortality. So when we focus on PM2.5, we should not discount the fact that PM10 is also heavily correlated with public health issues. However, Janssen et al. (2013) draws on data from 2008/2009, being aware of outdated measurements is of importance in our research.

Another gap in the research is the lack of structural measurement of ultrafine particles (UFP) in the Netherlands. These particles are smaller than PM2.5 and are also commonly associated with health risks. Some of the health risks that are commonly attributable to PM2.5 may partially have a causal relationship with UFP concentration, which is much more difficult to find consistent data for (CLO, 2024).

This is a relevant issue in Dutch society for a variety of reasons. For example, the debate around the role that TATA Steel plays when it comes to contributing to health risks for local citizens. For example, residents in Wijk aan Zee lose an average of 2.5 months of life expectancy due to exposure to particulate matter and nitrogen dioxide. Furthermore, approximately 4% of future lung cancer cases and 3% of childhood asthma cases in the immediate area are attributable to the plant's current emissions (Geelen et al., 2023).

Another reason this is relevant has to do with international guidelines. Currently, the Netherlands meets PM2.5 emission standards that are imposed by the European Union (EU). But the value that the World Health Organization (WHO) advises was exceeded in the whole country in 2023 (CLO, 2024b). Furthermore, the EU is set to lower its PM2.5 emission standards from 25 µg/m³ to 10 µg/m³ by the year 2030. Since these future EU standards are not yet met everywhere across the Netherlands, Dutch policy towards reducing PM2.5 emissions might have to become stricter in the upcoming years.

# Part 2 - Data Sourcing

## 2.1 Load in the data

Preferably from a URL, but if not, make sure to download the data and store it in a shared location that you can load the data in from. Do not store the data in a folder you include in the Github repository!

```{r loading_data}
library(readr)
concentratie_2024_PFS10_Haarlemmermeer <- read_delim("data/concentratie_2024_PFS10_Haarlemmermeer.csv", 
delim = ";", escape_double = FALSE, trim_ws = TRUE)
Part1 <- read_csv("data/DataExtract.csv")
Part2 <- read_csv("data/DataExtract2.csv")
combined_data <- rbind(Part1,Part2)
```

midwest is an example dataset included in the tidyverse package

## 2.2 Provide a short summary of the dataset(s)

```{r}
#head(dataset)
```

In this case we see 28 variables, but we miss some information on what units they are in. We also don't know anything about the year/moment in which this data has been captured.

```{r}

#summary(concentratie_2024_PFS10_Haarlemmermeer$`concentratie (µg/m³)`)
#concentratie_2024_PFS10_Haarlemmermeer$`concentratie (µg/m³)` <- as.numeric(concentratie_2024_PFS10_Haarlemmermeer$`concentratie (µg/m³)`)

#summary(concentratie_2024_PFS10_Haarlemmermeer$`concentratie (µg/m³)`)

opnamen <- read_delim("data/Ziekenhuisopnamen__diagnose_indeling_VTV__regio_02062026_164323.csv", 
                                                                            delim = ";", escape_double = FALSE, trim_ws = TRUE, 
                                                                            skip = 3)

```

These are things that are usually included in the metadata of the dataset. For your project, you need to provide us with the information from your metadata that we need to understand your dataset of choice.

## 2.3 Describe the type of variables included

Think of things like:

- Do the variables contain health information or SES information?

- Have they been measured by interviewing individuals or is the data coming from administrative sources?

*For the sake of this example, I will continue with the assignment...*

# Part 3 - Quantifying

## 3.1 Data cleaning

Say we want to include only larger distances (above 2) in our dataset, we can filter for this.

```{r data_cleaning}

cleaned_pollution_data<- combined_data
cleaned_pollution_data$code <- cleaned_pollution_data$`NUTS Code`

combined_data$code <- combined_data$`NUTS Code`
require(stringr)
require(tidyverse)
length(cleaned_pollution_data$code)
cleaned_pollution_data <- cleaned_pollution_data %>%
  dplyr::filter(str_length(code) == 5
)
cleaned_pollution_data <- cleaned_pollution_data %>%
  filter(
    Category                   == "Mortality",
    Outcome                    == "All causes",
    `Health Indicator`         == "Attributable deaths (AD)",
    Sex                        == "Total",
    `Description Of Age Group` == ">= 30 years of age",
    Scenario                   == "Baseline from WHO 2021 AQG"
  )

```

Please use a separate 'R block' of code for each type of cleaning. So, e.g. one for missing values, a new one for removing unnecessary variables etc.

## 3.2 Generate necessary variables

Variable 1

```{r}
#turning variable all-lowercase and into one word 
names(cleaned_pollution_data)[names(cleaned_pollution_data) == "Affected Population"] <- "affected_population"
```

Variable 2

```{r gen_var2}
#turning variable all-lowercase
names(cleaned_pollution_data)[names(cleaned_pollution_data) == "Year"] <- "year"
```

## 3.3 Visualize temporal variation

```{r}
#defining the axes
  name=cleaned_pollution_data$year 
  value=cleaned_pollution_data$`Air Pollution Population Weighted Average [ug/m3]`
  
#plotting graph, aestheticising 
require(ggplot2)

ggplot(cleaned_pollution_data, aes(x=year, y=`Air Pollution Population Weighted Average [ug/m3]`)) + 
  geom_bar(stat = "summary", fun = "mean",
           fill = "lightpink",
           colour = "purple") +
  
  geom_smooth(method = "lm", color = "blue", linewidth = 1) +
  
  labs(
    title = "AIR POLLUTION",
    y = "Air Pollution Population Weighted Average [ug/m3]",
    x = "Year"
  )+
  theme_minimal()
```

## 3.4 Visualize spatial variation

```{r}
#reading dataset, accounting for European delimination and skipping three rows
Ziekenhuisopnamen_diagnose_indeling_VTV_regio_02062026_164323 <- read_delim("data/Ziekenhuisopnamen__diagnose_indeling_VTV__regio_02062026_164323.csv", 
     delim = ";", escape_double = FALSE, trim_ws = TRUE, 
     skip = 3)
```

```{r visualise_map}
library(sf)
library(ggplot2)

# Load CBS GGD-regio map for 2023 (whole NL), WGS84
ggd <- st_read("https://cartomap.github.io/nl/wgs84/ggdregio_2023.geojson")

# Plot
ggplot(ggd) +
  geom_sf(fill = "lightblue", color = "white", linewidth = 0.2) +
  theme_void() +
  labs(title = "GGD-regio's Nederland")
```

```{r visualise_map}
library(sf)
library(dplyr)
library(stringr)
library(ggplot2)
library(readr)

# 1. Download the GGD region map
ggd_map <- st_read("https://cartomap.github.io/nl/wgs84/ggdregio_2023.geojson", quiet = TRUE)

# 2. Load the hospital admissions data safely without automatic column naming
raw_hospital_data <- read_delim("data/Ziekenhuisopnamen__diagnose_indeling_VTV__regio_02062026_164323.csv", 
                                delim = ";", col_names = FALSE, skip = 3)

# 3. Extract the GGD region names and the column for age-standardized Coronary Heart Disease
# Column 1 (X1) contains regions. Column 31 (X31) contains the target health indicator.
clean_health_data <- data.frame(
  raw_region     = raw_hospital_data$X1[9:33],  # Selects exactly the 25 GGD rows (omitting national total)
  admission_rate = raw_hospital_data$X31[9:33]  # Selects the matching column data
)

# 4. Convert text data to numbers and clean up the naming suffix
clean_health_data <- clean_health_data %>%
  mutate(
    # Fix Dutch decimals: replace "," with "." so R recognizes them as numbers
    admission_rate = as.numeric(str_replace(admission_rate, ",", ".")),
    
    # Strip away the trailing " (GG)" from text strings
    clean_region = str_replace(raw_region, " \\(GG\\)", "")
  )

# 5. Correct administrative text variations to guarantee a perfect map merge
clean_health_data <- clean_health_data %>%
  mutate(
    clean_region = case_when(
      str_detect(clean_region, "Gelderland-Midden") ~ "Veiligheids- en Gezondheidsregio Gelderland-Midden",
      str_detect(clean_region, "ZHZ") ~ "Dienst Gezondheid & Jeugd ZHZ",
      TRUE ~ clean_region
    )
  )

# 6. Merge map boundaries and data (Map object on the left side)
spatial_joined_ggd <- ggd_map %>%
  left_join(clean_health_data, by = c("statnaam" = "clean_region"))

# 7. Render the inverted spatial heatmap
ggplot(spatial_joined_ggd) +
  geom_sf(aes(fill = admission_rate), color = "white", linewidth = 0.2) +
  # 'direction = -1' flips the palette: low values are light, high values are dark
  scale_fill_viridis_c(option = "magma", name = "Admissions\nper 10k", direction = -1) +
  theme_void() +
  labs(
    title = "Spatial Variation of Cardiovascular Hospital Admissions",
    subtitle = "Age-standardized coronary heart disease admissions per 10,000 residents by GGD region",
    caption = "Source: CBS / VTV-Kompas"
  )
```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.5 Visualize sub-population variation

What is the poverty rate by state?

```{r visualise_subpopulations}

require(ggplot2)

ggplot(combined_data, aes(x = `Degree Of Urbanisation`, y = `Air Pollution Population Weighted Average [ug/m3]`)) +
  
geom_boxplot(
           fill = "lightpink",
           colour = "purple",
           alpha = 0.7) +
  coord_flip() +

  labs(
    title = "Distribution of Air Pollution Levels by Degree of Urbanisation",
    subtitle = "Comparing regional exposure across Dutch sub-populations",
    x = "Degree of Urbanisation",
    y = "Population-Weighted Air Pollution (ug/m3)"
  ) +
  theme_minimal() 
  

```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.6 Event analysis

Analyze the relationship between two variables.

```{r analysis}
cleaned_pollution_data <- cleaned_pollution_data %>%
  mutate(period = ifelse(year < 2020, "Before 2020", "2020 onwards"))

cleaned_pollution_data$period <- factor(
  cleaned_pollution_data$period,
  levels = c("Before 2020", "2020 onwards")
)
ggplot(cleaned_pollution_data, 
       aes(x = year, 
           y = `Air Pollution Population Weighted Average [ug/m3]`)) + 
  
  geom_bar(stat = "summary", fun = "mean",
           aes(fill = period),
           colour = "purple") +
  
  geom_smooth(aes(group = period, color = period),
              method = "lm", linewidth = 1, se = TRUE) +
  
  geom_vline(xintercept = 2019.5,
             linetype = "dashed",
             colour = "black",
             linewidth = 0.8) +
  
  annotate("text", x = 2019.7, 
           y = max(cleaned_pollution_data$`Air Pollution Population Weighted Average [ug/m3]`, 
                   na.rm = TRUE),
           label = "COVID-19 (2020)",
           hjust = 0, vjust = 1, size = 3.5, colour = "black") +
  
  scale_fill_manual(values = c("Before 2020" = "lightpink", 
                                "2020 onwards" = "aquamarine")) +
  scale_color_manual(values = c("Before 2020" = "lightblue", 
                                 "2020 onwards" = "darkblue")) +
  
  labs(
    title = "AIR POLLUTION — Pre vs. Post 2020",
    y = "Air Pollution Population Weighted Average [ug/m3]",
    x = "Year",
    fill = "Period",
    color = "Trend"
  ) +
  theme_minimal()
```

Relevant because: drop in pollution, less industrial activity, policy change? visible decrease

# Part 4 - Discussion

## 4.1 Discuss your findings

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the link to your PUBLIC repository here: ...

## 5.2 Reference list

Use APA referencing throughout your document.

1.  Janssen, N. et al., (2013). Short-term effects of PM2.5, PM10 and PM2.5–10 on daily mortality in the Netherlands. The Science of the Total Environment, 463–464, 20–26. <https://doi.org/10.1016/j.scitotenv.2013.05.062>

2.  Ultrafijnstof in de lucht. (2024, December 11). Compendium Voor De Leefomgeving. Retrieved June 8, 2026, from <https://www.clo.nl/indicatoren/nl062301-ultrafijnstof-in-de-lucht#referentie>

3.  Geelen, L., Bogers, R., Elberse, J., Houthuijs, D., Montforts, M., & Schuijff, M. (2023). De bijdrage van Tata Steel Nederland aan de gezondheidsrisico’s van omwonenden en de kwaliteit van hun leefomgeving. In Rivm (National Institute for Public Health and the Environment). <https://doi.org/10.21945/rivm-2023-0171>

4.  Fijnere fractie van fijn stof (PM2,5) in lucht, 2009-2023. (2024b, July 10). CLO. Retrieved June 8, 2026, from <https://www.clo.nl/indicatoren/nl053209-fijnere-fractie-van-fijn-stof-pm25-in-lucht-2009-2023#referentie>

5.  World Health Organization (WHO). (2006). Air quality guidelines for particulate matter, ozone, nitrogen dioxide and sulfur dioxide. <https://www.who.int/publications/i/item/WHO-SDE-PHE-OEH-06-02>

6. European Environment Agency. (2022). Air Quality Health Risk Assessments. https://www.eea.europa.eu/en/datahub/datahubitem-view/49930245-dc33-4c47-93b8-9512f0622ebc?activeAccordion=1084339

7. Centraal Bureau voor de Statistiek. (2025). Ziekenhuisopnamen; diagnose-indeling VTV, regio. In Centraal Bureau voor de Statistiek. https://www.cbs.nl/nl-nl/cijfers/detail/84523ned
```{r}
# 1. Laad benodigde packages
library(sf)
library(ggplot2)
library(dplyr)

# 2. Laad NUTS3-kaart van Europa en filter op Nederland
nuts3_all <- st_read("https://github.com/iiasa/scse-geojson/raw/main/nuts/NUTS_RG_03M_2024_4326.geojson")
nuts3_nl <- nuts3_all %>% filter(CNTR_CODE == "NL")

# 3. Bepaal het meest recente jaar in je data
recent_jaar <- max(combined_data$year, na.rm = TRUE)

# 4. Selecteer de juiste kolom (let op: exacte naam met 'm3')
data_heatmap <- combined_data %>%
  filter(Year == recent_jaar) %>%
  select(code, 
         concentration = `Air Pollution Population Weighted Average [ug/m3]`)

# 5. Koppel aan de kaart
nuts3_merged <- nuts3_nl %>%
  left_join(data_heatmap, by = c("NUTS_ID" = "code"))

# 6. Plot de spatial heatmap
ggplot(nuts3_merged) +
  geom_sf(aes(fill = concentration), color = "white", size = 0.2) +
  scale_fill_viridis_c(option = "plasma", 
                       na.value = "lightgrey",
                       name = "PM2.5 (µg/m³)") +
  labs(title = paste("Luchtvervuiling per NUTS3-regio,", recent_jaar),
       subtitle = "Populatiegewogen gemiddelde concentratie") +
  theme_void() +
  theme(legend.position = "right")
```

```{r visualise_map}
library(sf)
library(dplyr)
library(ggplot2)

nl_map <- st_read("https://cartomap.github.io/nl/wgs84/coropgebied_2023.geojson")

map_data_clean <- cleaned_pollution_data %>%
  filter(
    year == 2023, # Pick a single year to visualize on the map
  )

spatial_joined <- nl_map %>%
  left_join(map_data_clean, by = c("statnaam" = "NUTS Name"))

ggplot(spatial_joined) +
  geom_sf(aes(fill = `Air Pollution Population Weighted Average [ug/m3]`), color = "white", linewidth = 0.2) +
  scale_fill_viridis_c(option = "magma", name = "Pollution (ug/m3)", direction = -1) +
  theme_void() +
  labs(
    title = "Spatial Variation of Air Pollution in the Netherlands",
    subtitle = "Population-weighted PM2.5 averages by NUTS 3 regions (2023)"
  )
```

```{r analysis}
event_data <- cleaned_pollution_data %>%
  filter(
    year >= 2015
  ) %>%
  group_by(year) %>%
  summarise(avg_pollution = mean(`Air Pollution Population Weighted Average [ug/m3]`, na.rm = TRUE))

ggplot(event_data, aes(x = year, y = avg_pollution)) +
  geom_line(color = "purple", linewidth = 1.2) +
  geom_point(color = "purple", size = 3) +
  
  geom_vline(xintercept = 2020, linetype = "dashed", color = "darkred", linewidth = 1.2) +
  
  annotate("text", x = 2020.2, y = 12, label = "COVID-19 Lockdowns", color = "darkred ", fontface = "bold", hjust = 0) +
  
  labs(
    title = "Event Analysis: Impact of 2020 COVID-19 Lockdowns on Air Quality",
    subtitle = "Annual population-weighted PM2.5 averages across the Netherlands",
    x = "Year",
    y = "Average Air Pollution (ug/m3)"
  ) +
  scale_x_continuous(breaks = 2015:2023) + 
  theme_minimal()
```
