---
title: "Template"
author: "Studentnames and studentnumbers here"
date: "`r Sys.Date()`"
output: pdf_document
---

# Set-up your environment

```{r package_install, include=FALSE}
#install.packages("tidyverse")
#install.packages("yaml")
#install.packages("rmarkdown")
```

```{r packages}
require(tidyverse)
require(rmarkdown)
require(yaml)
```

# Title Page

Include your names

Include the tutorial group number

Include your tutorial lecturer's name

# Part 1 - Identify a Social Problem



## 1.1 Describe the Social Problem

Include the following:

- Why is this relevant?

- ...

# Part 2 - Data Sourcing

## 2.1 Load in the data

Preferably from a URL, but if not, make sure to download the data and store it in a shared location that you can load the data in from. Do not store the data in a folder you include in the Github repository!

```{r loading_data}
library(readr)
concentratie_2024_PFS10_Haarlemmermeer <- read_delim("data/concentratie_2024_PFS10_Haarlemmermeer.csv", 
delim = ";", escape_double = FALSE, trim_ws = TRUE)
Part1 <- read_csv("data/DataExtract.csv")
Part2 <- read_csv("data/DataExtract2.csv")
combined_data <- rbind(Part1,Part2)
```

midwest is an example dataset included in the tidyverse package

## 2.2 Provide a short summary of the dataset(s)

```{r}
#head(dataset)
```

In this case we see 28 variables, but we miss some information on what units they are in. We also don't know anything about the year/moment in which this data has been captured.

```{r}

#summary(concentratie_2024_PFS10_Haarlemmermeer$`concentratie (µg/m³)`)
#concentratie_2024_PFS10_Haarlemmermeer$`concentratie (µg/m³)` <- as.numeric(concentratie_2024_PFS10_Haarlemmermeer$`concentratie (µg/m³)`)

#summary(concentratie_2024_PFS10_Haarlemmermeer$`concentratie (µg/m³)`)

opnamen <- read_delim("data/Ziekenhuisopnamen__diagnose_indeling_VTV__regio_02062026_164323.csv", 
                                                                            delim = ";", escape_double = FALSE, trim_ws = TRUE, 
                                                                            skip = 3)

```

These are things that are usually included in the metadata of the dataset. For your project, you need to provide us with the information from your metadata that we need to understand your dataset of choice.

## 2.3 Describe the type of variables included

Think of things like:

- Do the variables contain health information or SES information?

- Have they been measured by interviewing individuals or is the data coming from administrative sources?

*For the sake of this example, I will continue with the assignment...*

# Part 3 - Quantifying

## 3.1 Data cleaning

Say we want to include only larger distances (above 2) in our dataset, we can filter for this.

```{r data_cleaning}

cleaned_pollution_data<- combined_data
cleaned_pollution_data$code <- cleaned_pollution_data$`NUTS Code`


combined_data$code <- combined_data$`NUTS Code`
require(stringr)
require(tidyverse)
length(cleaned_pollution_data$code)
cleaned_pollution_data <- cleaned_pollution_data %>%
  dplyr::filter(str_length(code) == 5
)
cleaned_pollution_data <- cleaned_pollution_data %>%
  filter(
    Category                   == "Mortality",
    Outcome                    == "All causes",
    `Health Indicator`         == "Attributable deaths (AD)",
    Sex                        == "Total",
    `Description Of Age Group` == ">= 30 years of age",
    Scenario                   == "Baseline from WHO 2021 AQG"
  )

```

Please use a separate 'R block' of code for each type of cleaning. So, e.g. one for missing values, a new one for removing unnecessary variables etc.

## 3.2 Generate necessary variables

Variable 1

```{r}
names(cleaned_pollution_data)[names(cleaned_pollution_data) == "Affected Population"] <- "affected_population"
```

Variable 2
```{r gen_var2}
names(cleaned_pollution_data)[names(cleaned_pollution_data) == "Year"] <- "year"
```

## 3.3 Visualize temporal variation

```{r}
# data <- cleaned_pollution_data(
  name=cleaned_pollution_data$year 
  value=cleaned_pollution_data$`Air Pollution Population Weighted Average [ug/m3]`
  

require(ggplot2)

# Barplot
ggplot(cleaned_pollution_data, aes(x=year, y=`Air Pollution Population Weighted Average [ug/m3]`)) + 
  geom_bar(stat = "summary", fun = "mean",
           fill = "lightpink",
           colour = "purple") +
  
  geom_smooth(method = "lm", color = "blue", linewidth = 1) +
  
  labs(
    title = "AIR POLLUTION",
    y = "Air Pollution Population Weighted Average [ug/m3]",
    x = "Year"
  )+
  theme_minimal()
```

## 3.4 Visualize spatial variation

```{r}
Ziekenhuisopnamen_diagnose_indeling_VTV_regio_02062026_164323 <- read_delim("data/Ziekenhuisopnamen__diagnose_indeling_VTV__regio_02062026_164323.csv", 
     delim = ";", escape_double = FALSE, trim_ws = TRUE, 
     skip = 3)
```

```{r visualise_map}
library(sf)
library(ggplot2)

# Load CBS GGD-regio map for 2023 (whole NL), WGS84
ggd <- st_read("https://cartomap.github.io/nl/wgs84/ggdregio_2023.geojson")

# Plot
ggplot(ggd) +
  geom_sf(fill = "lightblue", color = "white", linewidth = 0.2) +
  theme_void() +
  labs(title = "GGD-regio's Nederland")
```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.5 Visualize sub-population variation

What is the poverty rate by state?

```{r visualise_subpopulations}

require(ggplot2)

ggplot(combined_data, aes(x = `Degree Of Urbanisation`, y = `Air Pollution Population Weighted Average [ug/m3]`)) +
  
geom_boxplot(
           fill = "lightpink",
           colour = "purple",
           alpha = 0.7) +
  coord_flip() +

  labs(
    title = "Distribution of Air Pollution Levels by Degree of Urbanisation",
    subtitle = "Comparing regional exposure across Dutch sub-populations",
    x = "Degree of Urbanisation",
    y = "Population-Weighted Air Pollution (ug/m3)"
  ) +
  theme_minimal() 
  

```

Here you provide a description of why the plot above is relevant to your specific social problem.

## 3.6 Event analysis

Analyze the relationship between two variables.

```{r analysis}
#
```

Here you provide a description of why the plot above is relevant to your specific social problem.

# Part 4 - Discussion

## 4.1 Discuss your findings

# Part 5 - Reproducibility

## 5.1 Github repository link

Provide the link to your PUBLIC repository here: ...

## 5.2 Reference list

Use APA referencing throughout your document.


```{r}
# 1. Laad benodigde packages
library(sf)
library(ggplot2)
library(dplyr)

# 2. Laad NUTS3-kaart van Europa en filter op Nederland
nuts3_all <- st_read("https://github.com/iiasa/scse-geojson/raw/main/nuts/NUTS_RG_03M_2024_4326.geojson")
nuts3_nl <- nuts3_all %>% filter(CNTR_CODE == "NL")

# 3. Bepaal het meest recente jaar in je data
recent_jaar <- max(combined_data$year, na.rm = TRUE)

# 4. Selecteer de juiste kolom (let op: exacte naam met 'm3')
data_heatmap <- combined_data %>%
  filter(year == recent_jaar) %>%
  select(code, 
         concentration = `Air Pollution Population Weighted Average [ug/m3]`)

# 5. Koppel aan de kaart
nuts3_merged <- nuts3_nl %>%
  left_join(data_heatmap, by = c("NUTS_ID" = "code"))

# 6. Plot de spatial heatmap
ggplot(nuts3_merged) +
  geom_sf(aes(fill = concentration), color = "white", size = 0.2) +
  scale_fill_viridis_c(option = "plasma", 
                       na.value = "lightgrey",
                       name = "PM2.5 (µg/m³)") +
  labs(title = paste("Luchtvervuiling per NUTS3-regio,", recent_jaar),
       subtitle = "Populatiegewogen gemiddelde concentratie") +
  theme_void() +
  theme(legend.position = "right")
```

```{r}

```
